Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Bunched Type Theory

### 2024-01-11 paper/paper.tex
some notes from over break

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}

@@

\begin{document}
\title{Formal Grammars as Types in Non-commutative Bunched Type Theory}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

\begin{abstract}
  We propose the category of functions from strings to sets as a
  universal syntax-independent model of formal grammars, generalizing
  the set of strings model of formal language theory to account for
  ambiguity. The morphisms of this category give a notion of parse
  transformer, and isomorphism corresponds to \emph{strong
  equivalence} of grammars. We observe that the standard algebraic
  constructions of regular expressions and their extension to
  mu-regular expressions correspond to universal constructions in this
  category, which we axiomatize as a structure we call a 2-Kleene
  Algebra.
\end{abstract}

@@

Formal language theory alone is not sufficient as a specification of a
parser: a language $L \subseteq \Sigma^*$ is equivalently defined as
its indicator function $\chi_L : \Sigma^* \to \Prop$ mapping each
string $w$ to the proposition that it is in the language $w \in
L$. This can then serve as the specification for a language
\emph{recognizer}: a program $r$ of type $\Sigma^* \to \Bool$ such
that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$, but
recognition is insufficient for most tasks for which the theory of
parsing was developed: the production of \emph{semantic} structures
from synactic descriptions in linguistics, compilation or
deserialization. The output of a parser is not just a boolean but a
\emph{parse tree}\footnote{such parse trees are usually not
materialized in memory, as the construction of the parse tree is fused
with application of semantic actions, which can be seen as a kind of
fold over the generated tree.}.

@@

@@

\subsection{Syntax}

@@

Set types $A$ and set contexts $\Delta$ denotes sets, and
grammar types $\gamma$ and grammar contexts $\Gamma$ in context
$\Delta$ denote families of grammars $(\Set^{\Sigma^*})^\Delta$.  A

@@

Let $\phi$ be a propositional BI formula over $\Sigma$. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $\phi$:
\begin{enumerate}
\item On the one hand we have the semantics just defined, $\llbracket \phi \rrbracket w$
\item On the other hand, we can view the string $w$ as a formula
  $\lceil w \rceil$ and define a parse to be an equivalence classes of
  proofs $x:\lceil w \rceil \vdash \phi$
\end{enumerate}

It is not difficult to see that at least for the ``strictly positive''
formulae (those featuring only the positive connectives
$\vee,\bot,\epsilon,\otimes,\mu$ and the generators $c$) that every
element $t \in \llbracket \phi \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the language. However it is far less clear that every proof
$x:\lceil w \rceil \vdash p : \phi$ corresponds to some sequence of
introduction forms since proofs can include elimination forms as
well. So we want to prove a kind of \emph{canonicity} result for the
calculus, a generalization of the disjunction theorem of constructive
logic:

\begin{enumerate}
\item Every proof $\lceil w \rceil \vdash p : \phi \otimes \psi$ is equal to $(q,r)$ for some $q,r$
\item Every proof $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
\item Every proof $\lceil w \rceil \vdash p : \phi \vee \psi$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
\item Every proof $\lceil w \rceil \vdash p : c$ is equal to $x:c$
\item Every proof $\lceil w \rceil \vdash p : \mu X. \phi$ is equal to $\textrm{roll}(q)$ where $q : \phi(\mu X.\phi/X)$
\end{enumerate}

To prove this result we will use a simple Kripke logical relation.

@@

\section{Related Work}

TODO

%% Vaughan Pratt on the residual operator
\section{Future work}

@@

While parsing typically refers to the production of semantic objects
from their description as strings, many tasks in programming can be
viewed as parsing of more structured objects such as trees, trees with
binding structure or graphs. Fundamental to programming language
implementation is \emph{type checking}, analogous to language
recognition, or more generally \emph{typed elaboration}, analogous to
parsing, which produces a semantic object in addition to performing
some analysis.

>>>>

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

@@

\begin{document}
\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Bunched Type Theory}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

\begin{abstract}
  We propose a semantic framework for the study of formal
  grammars. First, we provide a syntax-independent notion of formal
  grammar as a function from strings to sets, generalizing the
  familiar notion of a language as a set of strings. The set of such
  grammars naturally form a very rich category whose notion of
  isomorphism corresponds to strong equivalence of grammars, and for
  which many language-theoretic constructs can be defined using
  universal properties.

  Based on these category-theoretic constructions we propose a new
  syntactic formalism for formal grammars: a two-level non-commutative
  bunched dependent type theory whose types can be interpreted as
  grammars, and terms as parse transformers. The non-dependent
  fragment of this type theory is already enough to express
  regular/context-free expressions as well as finite automata,
  generalizing the language theoretic formalisms such as Kleene
  algebras to incorporate language ambiguity. By incorporating a
  restricted form of type dependency, we can further express indexed
  grammars, pushdown automata and Turing machines. This provides a new
  form of ``logical characterization'' of grammar classes based on a
  substructural logic rather than ordinary first-order logic. Further,
  the type theory can not only express the grammars as types, but
  equivalences between grammars and parsers as terms of the grammar.

  We give this type theory a semantics in our category of grammars and
  prove a canonicity theorem that shows that every term in a context
  corresponding to a fixed string is equal to a term in a canonical
  form encoding a parse tree of that string.
\end{abstract}

@@

Formal language theory alone is not sufficient as a specification of a
parser: a language $L \subseteq \Sigma^*$ is equivalently defined as
its indicator function $\chi_L : \Sigma^* \to \Prop$ mapping each
string $w$ to the proposition that it is in the language $w \in
L$. This can then serve as the specification for a language
\emph{recognizer}: a program $r$ of type $\Sigma^* \to \Bool$ such
that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$ holds, but
recognition is insufficient for most tasks for which the theory of
parsing was developed: the production of \emph{semantic} structures
from synactic descriptions in linguistics, compilation or
deserialization. The output of a parser is not just a boolean but a
\emph{parse tree}\footnote{such parse trees are usually not
materialized in memory, as the construction of the parse tree is fused
with application of semantic actions, which can be seen as a kind of
fold over the generated tree.}.

@@

\paragraph{TODO: more stuff}

%% Vaughan Pratt on the residual operator

@@

\subsection{Syntax}
Judgments: $\Delta$ set ctx, $\Delta \vdash A$ set, $\Delta \vdash M :
A$ term, $\Delta \vdash \Gamma$ grammar ctx, $\Delta \vdash G$
grammar, $\Delta\pipe \Gamma \vdash p : G$ parse term.

@@

Propositional fragment: monoidal biclosed category that is also
bicartesian closed and has $\omega$-colimits. Call such a BIC.

Set theoretic fragment: category with families $\mathcal C$ + $\Pi,\Sigma,W$.

Combination: category with families $\mathcal C$ + a functor $L : C
\to \textrm{BIC}$.

Grammatical fragment: functor from $\mathcal C$ to the category of BI-models

Example: use $\Set$ as the CwF, use the functor $\Delta \mapsto
(\Set^{\Sigma^*})^\Delta$ as the model of the propositional fragment
using the pointwise-Day convolution monoidal structure.

@@

Let $\cdot \vdash G$ be a grammar. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $G$:
\begin{enumerate}
\item On the one hand we have the semantics just defined, $\llbracket G \rrbracket w$
\item On the other hand, we can view the string $w$ as a grammar
  $\lceil w \rceil$ and define a parse to be a parse terms $\cdot\pipe
  x:\lceil w \rceil \vdash p : G$ modulo $\beta\eta$ equivalence.
\end{enumerate}

It is not difficult to see that at least for the ``strictly positive''
formulae (those featuring only the positive connectives
$0,+,\epsilon,\otimes,\mu$ and the generators $c$) that every element
$t \in \llbracket G \rrbracket w$ is a kind of tree and that the nodes
of the tree correspond precisely to the introduction forms of the
type. However it is far less obvious that \emph{every} proof $x:\lceil
w \rceil \vdash p : \phi$ is equal to some sequence of introduction
forms since proofs can include elimination forms as well. To show that
this is indeed the case we give a \emph{canonicity} result for the
calculus, a generalization of the disjunction theorem of constructive
logic:
\begin{enumerate}
\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
\item There are no terms $\lceil w \rceil \vdash p : 0$
\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
  to $(M,q)$ where $\cdot \vdash M : A$
\end{enumerate}

To prove this result we will use a logical families construction by
constructing a second model. First, the category with families will be
the category of logical families over set contexts/types
$\Delta$/$A$. Then the propositional portion will be defined by
mapping a logical family $\hat \Gamma \to \Gamma$

@@

\section{Future work}

@@

While parsing typically refers to the production of semantic objects
from their description as strings, many tasks in programming can be
viewed as parsing of more structured objects such as trees, trees with
binding structure or graphs. Fundamental to programming language
implementation is \emph{type checking}, analogous to language
recognition, or more generally \emph{typed elaboration}, analogous to
parsing, which produces a semantic object in addition to performing
some analysis. A type system could then be interpreted as a \emph{tree
grammar}. Our definition of string grammars as functors from
$\Sigma^*$ to $\Set$ can then be easily adapted to functors to $\Set$
from the set of \emph{trees} freely generated form some signature of
nodes. Then just as the set of string grammars inherits a monoidal
structure from the free monoid of strings using the Day convolution,
this set of tree grammars inherits a kind of weak algebraic structure
corresponding to the tree constructors. This suggests an even more
unusual form of bunched type theory, where the monoidal concatenation
and unit are replaced with context constructors corresponding to the
tree constructors.
